---
title: "Problem Set 03: Data Wrangling"
author: "Cory Wolf"
date: last-modified
date-format: "[Last modified on] MMMM DD, YYYY HH:mm:ss zzz"
format: 
  html: default
  pdf: default
editor: source
---

```{r include = FALSE}
library(knitr)
library(tidyverse)
library(scales)
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", comment = NA, message = FALSE,  warning = FALSE)
```

In this problem set we will practice some of the key data manipulation tasks for describing, summarizing, and working with data. We will specifically review the following functions from the `dplyr` package:

-   `select`
-   `mutate`
-   `summarize`
-   `arrange`
-   `filter`
-   `group_by`


In addition we will review how to save objects using the `<-` assignment operator.

The following code loads the necessary packages for this problem set:

::: {.callout-caution icon="false" title="R Code"}
```{r}
library(ggplot2)
library(dplyr)
```
:::

# The Data

The following code chunk loads the data set `txhousing` and displays the data using `glimpse`:

::: {.callout-caution icon="false" title="R Code"}
```{r}
data(txhousing)
glimpse(txhousing)
#
#txhousing <- txhousing[sample(1:802),]
```
:::

These data are about housing in Texas. Each row is monthly data for a given city in Texas in a given year. There are multiple years of data for each city.

::: {.callout-note icon="false" title="Problem 1"}
Examine `txhousing` in the data viewer. You can accomplish this two different ways: A) click on the name of the data in the Environment pane, or b) type `View(txhousing)` in the **console**. What is the last city listed in the data set (in row 8602)? Use `R` code to programatically find the last city in `txhousing`.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 1 Answers"}
```{r}
# Type your code and comments inside the code chunk
txhousing %>%
  slice(n())
```

-   Wichita Falls
:::

::: {.callout-note icon="false" title="Problem 2"}
Examine the variable descriptions by typing `?txhousing` in the **console**. What is the `listings` variable in this data set?
:::

::: {.callout-important icon="false" collapse="false" title="Problem 2 Answers"}
-   Total active listings
:::

# Data Wrangling Review

## `select()`

Sometimes we want to pull out or extract just one or two columns of data. The following code will extract only the columns in the data set for the variables `sales` and `volume`.

```{r, eval = FALSE}
txhousing |> 
  select(sales, volume)
```

The `|>` symbol is called the **piping** operator. Here, it takes the `txhousing` **data frame** and "pipes" or feeds it into the `select` function. You can think of the `|>` symbol as the word "then".

Note that an assignment operator (`<-`) was not used in the code; consequently the selected values are not saved. In the following code, the results are saved in a data frame **ALSO** called `txhousing`. By putting `-` in front of the `date` variable `R` selects all variables **except** the `date` variable.

::: {.callout-caution icon="false" title="R Code"}
```{r}
txhousing <- txhousing |> 
  select(-date)
head(txhousing) 
```
:::

If you examine `txhousing` in the data viewer, the `date` variable is no longer included.

## `filter()`

The filter function allows you to pull out just the **rows** (cases or observations) you want, based on some criteria in **one of the columns**.

Imagine we wanted to reduce the data set to include data for only 2012 in the city of Austin. The code chunk below filters the `txhousing` to only include rows in which the year is 2012 **and** the city is Austin. The results are saved in a new data frame called `austin_12`.

::: {.callout-caution icon="false" title="R Code"}
```{r }
austin_12 <- txhousing |> 
  filter(year == 2012 & city == "Austin") 
  #Or filter(year == 2012, city == "Austin")
head(austin_12)
```
:::

::: callout-note
Note that we use `==` to identify the desired criteria.
:::

What if we wanted to restrict our data set to only years before 2004 and the City of Austin? Below we use the `<` symbol to accomplish this. Note we did not **SAVE** these results in a new data frame...so no new data frame showed up in our Environment pane, but the results print out immediately below the code chunk.

::: {.callout-caution icon="false" title="R Code"}
```{r}
txhousing |> 
  filter(year < 2004, city == "Austin") |> 
  head()
```
:::

What if we wanted to use multiple cities? Below we use the `|` symbol to indicate that the city could be Austin **OR** Abilene. In this case, we **saved** these results as a new data frame called `aust_ab` that appears in your Environment pane.

::: {.callout-caution icon="false" title="R Code"}
```{r}
aust_ab <- txhousing |> 
  filter(city == "Austin" | city == "Abilene")
head(aust_ab)
tail(aust_ab)
```
:::

## `mutate()`

The mutate function can add new columns (variables) to a data frame. For instance, the following will add a new column to the data called `vol_100k` that expresses volume in units of \$100,000.

::: {.callout-caution icon="false" title="R Code"}
```{r}
txhousing <- txhousing |>
  mutate(vol_100k = volume/100000)
head(txhousing)
```
:::

Note that we **SAVED** these results in new data frame called `txhousing`. This therefore **overwrote** the old `txhousing` data frame with a new version that contains this column. You can open the `txhousing` data frame in the viewer to confirm that it now contains this new column.

## `summarize()`

One of the first tasks in data analysis is often to get descriptive statistics that help to understand the central tendency and variability in the data. The `summarize()` command can take a column of data, and reduce it to a summary statistic.

For instance, the code below uses the `austin_12` data set made earlier to calculate the mean monthly number of `sales` in Austin in 2012.

```{r eval = FALSE}
austin_12 |> 
  summarize(x_bar_sales = mean(sales))
```

This code tells `R` to calculate the `mean` of the variable `sales`, and to save the results in a variable called `x_bar_sales`.

You can also calculate multiple summary statistics at once, and even for multiple variables. Below we also calculate a standard deviation `sd()` of `sales`, a minimum `min()` of the `volume` variable, a maximum `max()` of the `volume` variable, etc. The `n()` calculates sample size...or the number of rows/ cases in the data frame.

::: {.callout-caution icon="false" title="R Code"}
```{r}
austin_12 |> 
  summarize(x_bar_sales = mean(sales), 
            sd_sales = sd(sales), 
            min_vol = min(volume), 
            max_vol = max(volume), 
            mdn_list = median(listings), 
            iqr_list = IQR(listings),
            sample_size = n()) -> ans1
kable(ans1)
```
:::

Note that the names of the elements you calculate are user defined, like `xbar_sales`, `min_vol`, and `mdn_list`. You could customize these names as you like (but don't use spaces in your names).

## `arrange()`

You just determined that the maximum volume of monthly sales in Austin in 2012 was a total of `r dollar(ans1$max_vol)` ... but what if you wanted to know **WHAT MONTH** that occurred in?

::: {.callout-caution icon="false" title="R Code"}
```{r}
austin_12 |>
  arrange(desc(volume)) -> ans2
head(ans2, n = 3) |> 
  kable()
```
:::

The above code tells `R` to `arrange` the rows in the data set based on the `volume` column and to do so in `desc`ending order. Consequently, the row with the `r dollar(ans2$volume[1])` in sales is shown at the top. We can see that this `volume` occurred in the $`r ans2$month[1]`^{\text{th}}$ month (`r month.name[ans2$month[1]]`).

## `group_by()`

Sometimes we also want to calculate summary statistics across different levels of another variable. For instance, here we find the average number of monthly sales that occurred in Abilene and Austin across all years in the data set. Note that we **use the `aust_ab` data frame** we created earlier, to restrict our analysis to those two cities.

::: {.callout-caution icon="false" title="R Code"}
```{r}
aust_ab |> 
  group_by(city) |> 
  summarize(x_bar_sales = mean(sales)) -> results
results |> 
  kable()
```
:::

From the results we can see that there were an average of `r round(results$x_bar_sales[1],1)` sales per month in Abilene, and `r round(results$x_bar_sales[2],1)` sales per month in Austin.

We can give `R` multiple variables to group by. For instance, the following code returns the mean sales for each month in each city averaged across all the years.

::: {.callout-caution icon="false" title="R Code"}
```{r, label = "aus_ab"}
aust_ab |> group_by(city, month) |> 
  summarize(x_bar_sales = mean(sales)) -> results_ab
results_ab |> 
  head() |> 
  kable()
```
:::

The mean number of sales for Abilene in January (across all years) was `r round(results_ab$x_bar_sales[1],1)` homes.

# Independent Practice

## Basic Syntax

This first set of questions will help you practice basic syntax.

::: {.callout-note icon="false" title="Problem 3"}
Write a code chunk to remove the inventory variable. Save the results in a data frame called `txhousing`. Confirm the variable `inventory` has been removed.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 3 Answers"}
```{r}
# Type your code and comments inside the code chunk
txhousing <- txhousing %>%
  select(-inventory)

txhousing
```
:::

::: {.callout-note icon="false" title="Problem 4"}
Make a data set called `dallas_sub` that includes data only from the city of Dallas in 2012 and 2013. Verify that the dimensions of `dallas_sub` are 24 by 8.  Show the first six rows of `dallas_sub`.  Explain why the following code does not give the requested answer:
```{r, eval = FALSE}
txhousing |> 
  filter(city =="Dallas" & year == 2012 | year == 2013) |> dim()
```
:::

::: {.callout-important icon="false" collapse="false" title="Problem 4 Answers"}
```{r}
# Type your code and comments inside the code chunk

dallas_sub <- txhousing %>%
  filter(city == "Dallas", year %in% c(2012, 2013))

dim(dallas_sub)

head(dallas_sub)
```
:::

::: {.callout-note icon="false" title="Problem 5"}
Add a column to the `dallas_sub` data set called `prct_sold` that calculates the percentage of `listings` that were `sold` (`sales/listings * 100`). Be sure to **save** the results into a data frame called `dallas_sub`. Display the last six rows of `dallas_sub`.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 5 Answers"}
```{r}
# Type your code and comments inside the code chunk

dallas_sub <- dallas_sub %>% 
  mutate(prct_sold = (sales / listings) * 100)

tail(dallas_sub)
```
:::

::: {.callout-note icon="false" title="Problem 6"}
Calculate the **average** percentage of listings that were sold in Dallas **in each month across the years** based on your `dallas_sub` data set. Save the results of the calculation in a data frame called `dallas_summary` with the results stored in `mean_prct_sold`. Display the results of `dallas_summary`.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 6 Answers"}
```{r}

dallas_summary <- dallas_sub %>%
  group_by(month) %>%
  summarize(mean_prct_sold = mean(prct_sold, na.rm = TRUE))

dallas_summary
```
:::

::: {.callout-note icon="false" title="Problem 7"}
Arrange the `dallas_summary` in `desc`ending order based on the average percentage of listings that were sold in Dallas, so you can see **which month** had the greatest percentage of houses sold in Dallas on average from 2012-2013. You do not need to save the results.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 7 Answers"}
```{r}
# Type your code and comments inside the code chunk
dallas_summary %>%
  arrange(desc(mean_prct_sold))
```
:::

## More Advanced Wrangling

Please answer the following questions with text and/or code where appropriate. You may have to use multiple `dplyr` functions to answer each question. Think through the steps of how to get to the answer you are trying to find.

::: {.callout-note icon="false" title="Problem 8"}
Run the following code chunk. Study the code, and the output. Explain in your own words what this code chunk calculated. Specifically, compare this code to the code in Problems 4 through 7.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 8 Answers"}
```{r}
txhousing |> 
  filter(year == 2012 | year == 2013, city == "Dallas") |>
  mutate(prct_sold = sales/listings *100) |>
  group_by(month) |>
  summarize(mean_prct_sold = mean(prct_sold)) |> 
  arrange(desc(mean_prct_sold)) |> 
  kable()
```

-   Delete this and put your text answer here.
:::

::: {.callout-note icon="false" title="Problem 9"}
In January of 2015, what city had the fewest houses listed for sale? Report the city and the number of houses said city had listed for sale using inline `R` code.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 9 Answers"}
```{r}
# Type your code and comments inside the code chunk
txhousing %>%
  filter(year == 2015, month == 1) %>%
  group_by(city) %>%
  summarize(houses = n()) %>%
  arrange(houses)

```

-   The first row will give you the city with the fewest houses.
:::

::: {.callout-note icon="false" title="Problem 10"}
In 2012, in which month were the most houses sold in Texas? Report the month were the most houses sold in Texas and the number of houses sold for that month using inline `R` code.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 10 Answers"}
```{r}
# Type your code and comments inside the code chunk
txhousing %>%
  filter(year == 2012) %>%
  group_by(month) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  arrange(desc(total_sales))
```

-  Month 8
:::

::: {.callout-note icon="false" title="Problem 11"}
Generate a **single** table that shows the total number of houses sold in **Austin** in **2000 and 2001** (total over the entire period), & the total number of houses sold in **Dallas** in **2000 and 2001** (total over the entire period). This calculation requires a number of steps, so it might help you to first write out on paper the different steps you will need to take. That will help you set out a "blueprint" for tackling the problem. **Hint**: recall the `sum()` function can add values.
:::

::: {.callout-important icon="false" collapse="false" title="Problem 11 Answers"}
```{r}
# Type your code and comments inside the code chunk
txhousing %>%
  filter(year %in% c(2000, 2001),
         city %in% c("Austin", "Dallas")) %>%
  group_by(city) %>%
  summarize(total_houses_sold = sum(sales, na.rm = TRUE))
```
:::

# Turning in Your Work

You will need to make sure you commit and push all of your changes to the github education repository where you obtained the lab.

::: callout-tip
-   Make sure you **render a final copy with all your changes** and work.
-   Look at your final html file to make sure it contains the work you expect and is formatted properly.
:::

# Logging out of the Server

There are many statistics classes and students using the Server. To keep the server running as fast as possible, it is best to sign out when you are done. To do so, follow all the same steps for closing Quarto document:

::: callout-tip
-   Save all your work.
-   Click on the orange button in the far right corner of the screen to quit `R`
-   Choose **don't save** for the **Workspace image**
-   When the browser refreshes, you can click on the sign out next to your name in the top right.
-   You are signed out.
:::

```{r}
sessionInfo()
```
